Measurement linkage using Markov Link Intervals

### 2019-05-10 paper/paper.tex
New structure!

\title{Unpaired measurement integration using Markov Link Intervals}

@@

\begin{abstract}
We consider two datasets, each measuring a different of observations on a common population.  The first dataset is constructed by drawing individuals randomly from this population and measuring a quantity about each individual.   The second dataset is constructed similarly, but a different quantity is measured.  ``Unpaired measurement integration'' attempts to use this data to infer the joint distribution of the two different quantities.  We overcome the unidentifiability problems in this task by measuring an additional quantity, similar to an instrumental variable.  This quantity is observed in both datasets and satisfies a conditional independence assumption.  We characterize identifiability in the presence of these variables and develop a consistent estimator.  The performance of the estimator is tested in simulations.  The method is applied to transcriptomic measurements to uncover the relationship between two different technologies for measuring gene expression in mouse neurons.
\end{abstract}

@@

Unpaired measurement integration seeks to infer the joint distribution on two variables $(X,Y)$ using knowledge of the marginal distributions on $X$ and $Y$.  This problem is ill-posed, but practical algorithms have nonetheless been developed.  An early approach by Andrey Kolmogorov noted that the expectation for any linear function $f(X,Y)=\alpha X+\beta Y$ could be effectively bounded using only the marginal distributions \cite{makarov1982estimates}.  Bounds on arbitrary univariate functions $f(X,Y)$ have also been developed in the context of causal inference \cite{balke1994counterfactual,FanYan2017}.  More recently, the CycleGan algorithm has gained prominence in the computer graphics community for succesfully estimating a joint distribution between photographs ($X$) and watercolor paintings ($Y$) of the same subject; CycleGan accomplishes this unpaired measurement integration using two completely separate datasets, one of photographs and one of watercolor paintings \cite{zhu2017unpaired}.

To resolve the fundamental ill-posedness of unpaired dataset integration, some form of extra information is often employed.  We assume this side-information variable can be observed for each sample in both datasets.  The work of Fan \emph{et al} (2017) gives a detailed overview for this case \cite{FanYan2017}.  Fan characterizes the asymptotics of estimators for $\mathbb{E}[f(X,Y)]$, where $f$ can takes a wide variety of forms.  Estimator asymptotics are established under many different assumptions on the relationship between the measurements and the side-information

Inspired by problems in the study of cell populations, this paper considers a new kind of assumption about the side-information.  Let $L$ denote the side-information.  Suppose that $L$ and $Y$ are conditionally independent given $X$.  This assumption holds naturally whenever both $L$ and $Y$ can both be understood as independent noisy measurements of $X$.  We give a complete real-world example in Section \ref{sec:allen}.  The variable $L$ bears a formal similarity to an instrumental variable, but its function is different: instrumental variables resolve causal effects in the presence of unobserved confounders, whereas these variables permit joint inference when only marginal observations are available.

Inference under the conditional independence assumption must overcome two obstacles: the datasets only allow approximate estimation for the marginal distributions, and the marginal distributions only give a partial picture of the joint.  These problems cannot be considered separately.  If we treat statistical estimates as exact truth and proceed with a purely algebraic analysis of the model, inference can go badly wrong.  It turns out that even infinitesimal estimation errors can result in substantially incorrect conclusions, as we show in Lemma \ref{lem:noobseq}.   This marks a stark contrast from the conditions considered in Fan \emph{et al} (2017), where additional statistical noise results in a graceful degradation of inference accuracy.

In this paper, we overcome both of these challanges for the discrete setting.  When $L,X,Y$ are categorical distributions it is possible to make a detailed study of what are known as the observational equivalence classes \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.  Each class corresponds to a set of parameters that give rise to the same distribution on the data we observe.  We can achieve consistent inference by taking a carefully chosen dilation of these sets.  We design a simple optimization algorithm to find extremal points in these dilated sets.  These extremal points yield bounds for the parameters which we dub ``Markov Link Intervals,'' because they use the Markov conditional independence assumption to bound the parameter that links $X$ to $Y$.    Performance is investigated through a variety of simulation studies.  Finally, the estimator is applied to uncover the relationship between two different technologies for measuring gene expression in mouse neurons.

@@

\section{Formal problem statement and notation}

@@

Suppose we are given independent samples from $p$, but some data is missing.  Each sample includes $L,X$ or $L,Y$ but no sample ever includes both $X$ and $Y$.  Can we still learn $p$?  Let $n,m>0$ and $(L_1,X_1,Y_1),(L_2,X_2,Y_2) \cdots (L_{n+m},X_{n+m},Y_{n+m})$ denote samples of independent copies of $(L,X,Y)$.  Let
%
\begin{align*}
\DX &= (L_1,X_1),\cdots (L_n,X_n) \\
\DY &= (L_{n+1},Y_{n+1}),\cdots (L_{n+m},Y_{n+m}) 
\end{align*}
%
Our task is to use the data $(\DX,\DY)$ to estimate the parameters of $p$, namely $(\pi,\alpha,\beta)$.  Estimating $\pi,\alpha$ from this data is fairly straightforward.  However, existing estimation theory is less helpful for the parameter $\beta$.  In this paper we develop theory and practice for estimating this parameter.

@@

\section{Markov Link Intervals (MLI) for estimating $\beta$}

@@

\begin{lemma}[Unidentifiability]  \label{lem:unident} When $N_L<N_X$ there is no generally consistent estimator to estimate $\beta$ from the data $(\DX,\DY)$.  
\end{lemma}

@@

A more general solution can be obtained by seeking a small dilation of the observational equivalence classes.  Let $\LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta)$ denote the likelihood of the observed data under the model under the parameters $\hat \pi,\hat \alpha,\hat \beta$.  Let $\LL^*$ denote the maximum value of $\LL$ over all possible parameters.  Define
\[
\hat \Theta=\left\{\hat \pi, \hat \alpha,\hat \beta:\ \LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta) > \LL^* - \log(n+m)\right\}
\]
This log-scaled likelihood envelope has two desirable properties: it both concentrates around the true distribution on the observables and is asymptotically guaranteed to contain the true model parameters.
\begin{thm}[Consistency of $\hat \Theta$]\label{thm:consistency} Let $n,m\rightarrow\infty$ such that $|\log(n/m)|<M$.  Then:
\begin{enumerate}
    \item $\hat \Theta$ is asymptotically guaranteed to contain the true parameter values.  
    \item $\hat \Theta$ concentrates for $\pi,\alpha,\alpha\beta$. Let $\xi(\pi',\alpha',\beta')$ denote the maximum of the Euclidean differences $|\pi-\pi'|$, $|\alpha-\alpha'|$, $|\alpha\beta-\alpha'\beta'|$.  Then $\max_{\theta \in \hat \Theta} \xi(\theta)$ converges to zero in probability.   
\end{enumerate}
\end{thm}
This theorem indicates a practical solution for estimating $\beta$.  For each $x,y$, we define the \emph{Markov Link Interval} as 
\[
[\hat \beta^-_{x,y},\hat \beta^+_{x,y}] \triangleq \left[\min_{\pi',\alpha',\beta' \in \hat \Theta} \beta'_{x,y},\max_{\pi',\alpha',\beta' \in \hat \Theta}\beta'_{x,y}\right]
\]  
Theorem \ref{thm:consistency} shows that $\beta_{x,y}$ is asymptotically included in the interval $[\hat \beta^-_{x,y},\hat \beta^+_{x,y}]$.

To calculate the Markov Link Interval we must solve the associated optimization problems.  Due to the high dimensionality of the problem and the ill-conditioning of the matrices involved, we were unable to tune off-the-shelf optimization methods to reliably solve these problems.  We developed an alternative optimization procedure which relies on a minorization-maximization approach and requires no tuning parameters.  Appendix \ref{sec:mlmdetails} provides a detailed outline of the algorithm.\footnote{Code is published at https://github.com/jacksonloper/markov-link-method, including a tutorial ipython notebook that details every computation made in this paper.}

@@

\begin{itemize}
    \item Varying the sample size and the strength of the $L,X$ dependency.  Fix $N_X=N_Y=20$ and $N_L=15$.  We perform a series of trials.  In each trial we select $\pi$ and the rows of $\beta$ uniformly from the simplex.  The rows of $\alpha$ are chosen more carefully.  We will first generate $\tilde \alpha_1$ by sampling the rows uniformly from the simplex.  We then generate $\tau$ uniformly from the simplex and define $\tilde \alpha_2 = \mathbf{1}\tau^T$, i.e. every row of $\alpha_2$ is given by $\tau$.  Finally, we select $c\in[0,1]$ and define $\alpha=(1-c)(\tilde \alpha_1)+c\tilde \alpha_2$.  By varying $c$, we can control the mutual information between $L$ and $X$.  When $c=0$ the variable $L$ gives significant information about $X$; when $c=1$ we have that $L$ is independent of $X$.  Once $\pi,\alpha,\beta$ are selected, we select a sample size $n=m$, simulate a dataset, compute a maximum likelihood estimator $\hat \beta $ (see Appendix \ref{sec:mlmdetails} for our exact algorithm), and compute the average variation distance between $\beta$ and our estimate $\hat \beta$.  This average total variation distance is computed by averaging the row-by-row total variations between $\beta$ and  $\hat \beta$.  After performing a series of such trials, we visualize the results to obtain qualitative understanding of the performance of the maximium likelihood estimator $\hat \beta$.

@@

Figure \ref{fig:ident} shows the results.  We see that $\beta$ is much harder to estimate in some situations.  For example, the first simulation (Figure \ref{fig:ident}a) demonstrates that $\beta$ is harder to estimate when $L,X$ have low mutual information.  The second simulation (Figure \ref{fig:ident}b) shows a more complex picture; in some cases it is actually possible to get perfect estimation of $\beta$, even though $N_L<N_X$.  If $X,Y$ are independent or deterministically related, we see that perfect estimation becomes possible in the limit of infinite data.  However, in the more typical case where the relatioship between $X$ and $Y$ neither independent nor deterministic, it appears that consistent estimation is not possible.

@@

\hfill{}\begin{minipage}{.9\textwidth}
\caption{\textbf{Cell-type science.  Maximum likelihood estimation of $\beta$ suggests unlikely results about cell biology; Markov Link Intervals indicate plausible bounds.}  \label{fig:allen}
%
We performed unpaired measurement integration for two ways of measuring the type of cells from mouse visual cortex.  The first method uses gene expression to obtain a definitive measurement of the cell-type, $X$.  The second method yields a considerably less reliable estimate for the cell-type, $Y$.  For each cell we also know the value of $L$, which corresponds to another noisy indicator about the cell-type $X$.  In one dataset we can observe $(L,X)$ for many cells; in another dataset we can observe $(L,Y)$ many other cells.  Using this data we construct three estimates for $\beta$, the matrix of parameters which defines the relationship between $X$ and $Y$.  For each possible index $(x,y)$, $\hat \beta^{-}_{x,y}$ estimates a lower-bound estimate for $\beta_{x,y}$, $\hat \beta_{x,y}$ is a maximum likelihood estimator, and $\hat \beta^{+}$ estimates an upper bound.  Due to identifiability issues $\hat \beta_{x,y}$ is not guaranteed to be consistent.  However, Theorem \ref{thm:consistency} shows that the Markov Link Interval $[\hat \beta^{-}_{x,y},\hat \beta^{+}_{x,y}]$ is asymptotically guaranteed to include the truth.  Above we show all three of these estimators for a selection of values of $(x,y)$; we visualize these values in a grid, where rows correspond to values of $x$, columns correspond to values of $y$, and the color of the square in the $(x,y)$ entry indicates the value of the corresponding parameter.  The values of $\hat \beta$ are nonsensical in many cases.  On the other hand, the Markov Link Interval suggests plausible bounds for the parameters and also indicates further experiments which would help refine these bounds.
}
\end{minipage}\hfill{}
\end{minipage}}\end{figure}

@@

For any cell it is thus impossible to measure a rich variety of cell properties \emph{and} reliably estimate the cell-type.  To understand the relationship between cell-type and these other cell properties, we therefore need some form of unpaired measurement integration.  This integration would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.  To facilitate this integration, one can conduct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.  In each experiment, the probability that a particular cell will be picked out is entirely determined by the cell type $X$, and is thus statistically isolated from the noisy cell-type estimate $Y$.  By varying the experimental parameters we can change the distribution on which cells are picked out.  Letting $L$ denote these corresponding experimental parameters, we see that $L$ can act as a noisy independent indicator about the cell-type.  Markov Link Intervals should therefore give asymptotically consistent bounds on the reliability of the noisy estimate $Y$.

@@

The Markov Link Intervals yield plausible bounds on the true model parameters.  For example, the lower bounds of the Markov Link Interval suggests that a sizeable fraction of `Lamp5 Pdlim5' cells are indeed classified as `Lamp5' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as `Lamp5 Pdlim5', but instead only gives coarse class-level cell-type predictions, such as `Lamp5').  The upper bounds show that these cells are not commonly categorized into any group besides the `Lamp5' type and the the `unknown' type which the Patch method assigns to cells which it cannot identify.  Consider another example: for cells of type `Sst Myh8-1',  Markov Link Intervals give us considerably less certainty regarding the related type.  The intervals indicate that the Patch method does not typically classify these cells as `CR,' `L6b,' `Meis2,' `Sncg,' or `Vip' -- but it does not indicate how they are being classified.  At the extreme end, we have cells of type `Lamp5 Elgn3-1,' for which the Markov Link Intervals are completely vacuous.   It is possible that the Patch method perfectly classifies these `Lamp5 Elgn3-1' cells; it is possible that it always misclassifies them; the available data cannot tell us either way.

Parameters with large intervals suggest new experiments to perform in order to more closely determine the model parameters.  For example, the significant parameter ambiguity for cells with type `Lamp5 Egln 1' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.  If we could find a Cre/Lox procedure that obtained many `Lamp5 Egln-1' cells but no cells which Patch categorizes as `Pvalb,' this would show that `Lamp5 Egln 1' cells are not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells and also selects cells which are largely categorized as `Pvalb' by Patch, this would show the opposite.  If these additional experiments were performed, new Markov Link Intervals could be computed on the new data to determine what aspects of the link are still ambiguous.

@@

We here study `unpaired measurement integration,' the problem of estimating a joint distribution $(X,Y)$ from two separate datasets, one with samples of $X$ and another with samples of $Y$.  Rigorous statistical analysis of this ill-posed problem are complicated by statistical error and identifiability issues.  Here we characterize both issues in the discrete settings when we can observe a third variable $L$ for each sample in both dataset, under the assumption that $L$ and $Y$ are independent conditioned in $X$.  This assumption arises naturally if $L$ gives noisy measurement about $X$ and that noise is independent of $Y$.  Under this assumption, we construct ``Markov Link Intervals'' which give provably consistent bounds on the joint distribution.  Simulations bear out the consistency of these bounds and also suggest estimation is easiest performs best when $L$ is predictive of $X$ and the relationship between $X$ and $Y$ is either deterministic or completely independent.  An application of the method to cell-type populations demonstrates the method's practical utility for unpaired measurement integration problems.  One can imagine that these results could be extended to the continuous setting, though additional regularity conditions would be required.

@@

@@

\begin{lemma*}[Unidentifiability]  When $N_L<N_X$ there is no generally consistent estimator to estimate $\beta$ from the data $(\DX,\DY)$. 
\end{lemma*}

>>>>

\title{Measurement linkage using Markov Link Intervals}

@@

\begin{abstract}
We consider two datasets, each taking different kinds of measurements from samples of a common population.  The first dataset is constructed by drawing individuals randomly from this population and measuring a quantity about each individual.   The second dataset is constructed similarly, but a different quantity is measured.  ``Unpaired measurement linkage'' attempts to use this data to infer the joint distribution of the two different quantities.  We study this problem in the presence of a third quantity which is observed in both datasets and satisfies a conditional independence assumption.  We characterize identifiability in this case and develop estimators for the joint distribution.  The performance of the estimators is tested in simulations.  The method is applied to transcriptomic measurements to uncover the relationship between two different technologies for measuring gene expression in mouse neurons.
\end{abstract}

@@

Many scientific experiments produce data that come from different measurement techniques.  To use such data, we need to understand the link between the different types of measurements.  For example, in a biological context, a transcriptomic-morphological link would answer a question like:  ``We have measured the transcriptome of this cell -- what might its morphology have been?''

It is particularly challenging to estimate the link when multiple measurements of the same specimen are unavailable.  For example, it is currently often difficult to observe both the RNA expression and morphological properties of a single cell.  Rather, we typically have two datasets: one contains the transcriptomes of one group of cells and the other contains the morphology of another group of cells from the same population.   However, the datasets are not completely disconnected: there is some information which is collected for every specimen in both datasets.  For example, it may be possible to read optogenetic marker information for every cell in both datasets.  The task is to leverage this information to uncover the link.

This \emph{unpaired measurement linkage} problem arises in many scientific and technological fields where it is difficult to take multiple types of measurements of the same specimen.  Formally, unpaired measurement linkage seeks to estimate a distribution $p(\ell,x,y)$ even though we can never observe a complete sample $(L,X,Y)\sim p$.  For every sample we can observe $(L,X)$ or we can observe $(L,Y)$ -- but we can never observe $(X,Y)$.  In the cell biology example above, $X$ might correspond to a cell's transcriptome, $Y$ might correspond to the cell's morphology, and $L$ might correspond to an optogenetic marker that can be observed for every cell.

The key challenge in unpaired measurement linkage is to identify good assumptions or inductive biases; without any additional assumptions the problem is always ill-posed.  An overview of different possible assumptions can be found in Fan \emph{et al} \cite{FanYan2017}.

In this paper, we seek to understand unpaired measurement linkage in the discrete setting under the assumption that $L$ and $Y$ are conditionally independent given $X$.  This assumption is inspired by cell biology, and holds naturally whenever $L$ and $Y$ can be understood as independent noisy measurements of $X$.  For example, in Section \ref{sec:allen} we will see an example where optogenetic markers ($L$) and a low-fidelity transcriptomic record ($Y$) act as independent noisy indicators about cell-type ($X$).  Many other examples can be found in Section \ref{sec:conclusions}.  Note that the side-information $L$ bears a formal similarity to an instrumental variable, but its function is different; instrumental variables resolve causal effects in the presence of unobserved confounders, whereas these variables permit joint inference when only marginal observations are available.

We investigate the identifiability issues and develop estimators which are asymptotically guaranteed to bound the true joint distribution.  The key idea is to make a detailed study of the ``observational equivalence classes'' \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.  Each class corresponds to a set of parameters that give rise to the same distribution on the data we observe.  We show that estimating these sets is actually impossible, but a carefully-chosen dilation of these sets asymptotically contains the true model parameters and also concentrates on the observable distributions.  We also design a simple optimization algorithm to find extremal points in these dilated sets.  These extremal points yield bounds for the parameters which we dub ``Markov Link Intervals,'' because they use the Markov conditional independence assumption to bound the parameter that links $X$ to $Y$.   Finally, we interrogate the effectiveness of these intervals using simulations and real-world data.

This work arises out of a literature that spans many fields and many decades.  Perhaps the earliest approach to unpaired measurement integration comes from Andrey Kolmogorov, who conjectured an inequality for the cumulative distribution function of the sum of two random variables in terms of their marginal distributions \cite{makarov1982estimates}.  Related inequalities have also been studied in the context of quantum physics \cite{chaves2014inferring}.  More recently, bounds on arbitrary univariate functions $f(X,Y)$ have also been developed in the context of causal inference \cite{balke1994counterfactual}.  Most recently, the CycleGan algorithm has gained prominence in the computer graphics community for successfully estimating a joint distribution between photographs ($X$) and watercolor paintings ($Y$) of the same subject; CycleGan accomplishes this unpaired measurement linkage using two completely separate datasets, one of photographs and one of watercolor paintings \cite{zhu2017unpaired}.  This work connects most directly to the work of Fan \emph{et al} (2017), which gives an extensive study for unpaired measurement linkage problem under many different assumptions \cite{FanYan2017}.  Under each assumption the asymptotics of estimators for $\mathbb{E}[f(X,Y)]$ are characterized, where $f$ can takes a wide variety of forms.  Here we extend this work by considering similar questions under a new kind of assumption.

@@

\section{Markov Link Intervals (MLI)}

@@

Suppose we are given independent samples from $p$, but some data is missing.  Each sample includes $L,X$ or $L,Y$ but no sample ever includes both $X$ and $Y$.  Can we still learn $p$?  Let $n,m>0$ and $(L_1,X_1,Y_1),(L_2,X_2,Y_2) \cdots (L_{n+m},X_{n+m},Y_{n+m})$ denote samples of independent copies of $(L,X,Y)$.  Let
%
\begin{align*}
\DX &= (L_1,X_1),\cdots (L_n,X_n) \\
\DY &= (L_{n+1},Y_{n+1}),\cdots (L_{n+m},Y_{n+m}) 
\end{align*}
%
Unpaired measurement linkage seeks to use the data $(\DX,\DY)$ to estimate the parameters of $p$, namely $(\pi,\alpha,\beta)$.  Observe that estimating $\pi,\alpha$ from this data is straightforward.

Existing estimation theory is less helpful for the parameter $\beta$.  In this paper we develop theory and practice for estimating $\beta$.

@@

\subsection{Theory}

@@

\begin{lemma}[Unidentifiability]  \label{lem:unident} When $N_L<N_X$ there is no generally consistent estimator to infer $\beta$ from the data $(\DX,\DY)$.  
\end{lemma}

@@

A more general solution can be obtained by seeking a small dilation of the observational equivalence classes.  Let $\LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta)$ denote the likelihood of the observed data under the model under the parameters $\hat \pi,\hat \alpha,\hat \beta$.  Let $\LL^*$ denote the maximum value of $\LL$ over all possible parameters.  Define
\begin{gather}
\hat \Theta=\left\{\hat \pi, \hat \alpha,\hat \beta:\ \LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta) > \LL^* - \log(n+m)\right\} \label{eq:magictheta}
\end{gather}
This log-scaled likelihood envelope has two desirable properties: it both concentrates around the true distribution on the observables and is asymptotically guaranteed to contain the true model parameters.
\begin{thm}[Consistency of $\hat \Theta$]\label{thm:consistency} Let $n,m\rightarrow\infty$ such that $|\log(n/m)|<M$.  Then:
\begin{enumerate}
    \item $\hat \Theta$ is asymptotically guaranteed to contain the true parameter values.  
    \item $\hat \Theta$ concentrates for $\pi,\alpha,\alpha\beta$. Let $\xi(\pi',\alpha',\beta')$ denote the maximum of the Euclidean differences $|\pi-\pi'|$, $|\alpha-\alpha'|$, $|\alpha\beta-\alpha'\beta'|$.  Then $\max_{\theta \in \hat \Theta} \xi(\theta)$ converges to zero in probability.   
\end{enumerate}
\end{thm}

\subsection{Practice}

To estimate the parameters $\beta_{x,y} = p(y|x)$, we propose \emph{Markov Link Intervals}.  For each $x,y$ we define this interval as 
\[
[\hat \beta^-_{x,y},\hat \beta^+_{x,y}] \triangleq \left[\min_{\pi',\alpha',\beta' \in \hat \Theta} \beta'_{x,y},\max_{\pi',\alpha',\beta' \in \hat \Theta}\beta'_{x,y}\right]
\]  
where, as above, $\hat \Theta$ is the set of parameters whose likelihood is within $\log n+m$ of the maximum likelihood.  Theorem \ref{thm:consistency} shows that $\beta_{x,y}$ is asymptotically included in the interval $[\hat \beta^-_{x,y},\hat \beta^+_{x,y}]$.

To calculate the Markov Link Intervals from data, we must solve the associated constrained optimization problems.  Due to the high dimensionality of the problem and the ill-conditioning of the matrices involved, we were unable to tune off-the-shelf optimization methods to reliably solve these problems.  We developed an alternative optimization procedure which requires no tuning parameters.  The key idea is that the Lagrangian of this problem can itself be understand as a penalized likelihood; for any fixed value of the Lagrange multiplier, the likelihood can be thus be optimized using an Expectation Maximization algorithm.  The correct value of the Lagrange multiplier can be found with a simple line-search.  Appendix \ref{sec:mlmdetails} provides a detailed outline of the algorithm.\footnote{Code is published at https://github.com/jacksonloper/markov-link-method, including a tutorial ipython notebook that details every computation made in this paper.}

@@

\begin{itemize}
    \item Varying the sample size and the strength of the $L,X$ dependency.  Fix $N_X=N_Y=20$ and $N_L=15$.  We perform a series of trials.  In each trial we select $\pi$ and the rows of $\beta$ uniformly from the simplex.  The rows of $\alpha$ are chosen more carefully.  We will first generate $\tilde \alpha_1$ by sampling the rows uniformly from the simplex.  We then generate $\tau$ uniformly from the simplex and define $\tilde \alpha_2 = \mathbf{1}\tau^T$, i.e. every row of $\alpha_2$ is given by $\tau$.  Finally, we select $c\in[0,1]$ and define $\alpha=(1-c)(\tilde \alpha_1)+c\tilde \alpha_2$.  By varying $c$, we can control the mutual information between $L$ and $X$.  When $c=0$ the variable $L$ gives significant information about $X$; when $c=1$ we have that $L$ is independent of $X$.  Once $\pi,\alpha,\beta$ are selected, we select a sample size $n=m$, simulate a dataset, compute a maximum likelihood estimator $\hat \beta $ (see Appendix \ref{sec:mlmdetails} for our exact algorithm), and compute the average variation distance between $\beta$ and our estimate $\hat \beta$.  This average total variation distance is computed by averaging the row-by-row total variations between $\beta$ and  $\hat \beta$.  After performing a series of such trials, we visualize the results to obtain qualitative understanding of the performance of the maximum likelihood estimator $\hat \beta$.

@@

Figure \ref{fig:ident} shows the results.  We see that $\beta$ is much harder to estimate in some situations.  For example, the first simulation (Figure \ref{fig:ident}a) demonstrates that $\beta$ is harder to estimate when $L,X$ have low mutual information.  The second simulation (Figure \ref{fig:ident}b) shows a more complex picture; in some cases it is actually possible to get perfect estimation of $\beta$, even though $N_L<N_X$.  If $X,Y$ are independent or deterministically related, we see that perfect estimation becomes possible in the limit of infinite data.  However, in the more typical case where the relationship between $X$ and $Y$ neither independent nor deterministic, it appears that consistent estimation is not possible.

@@

\hfill{}\begin{minipage}{.9\textwidth}
\caption{\textbf{Cell-type science.  Maximum likelihood estimation of $\beta$ suggests unlikely results about cell biology; Markov Link Intervals indicate plausible bounds.}  \label{fig:allen}
%
We performed unpaired measurement linkage for two ways of measuring the type of cells from mouse visual cortex.  The first method uses gene expression to obtain a definitive measurement of the cell-type, $X$.  The second method yields a considerably less reliable estimate for the cell-type, $Y$.  For each cell we also know the value of $L$, which corresponds to another noisy indicator about the cell-type $X$.  In one dataset we can observe $(L,X)$ for many cells; in another dataset we can observe $(L,Y)$ many other cells.  Using this data we construct three estimates for $\beta$, the matrix of parameters which defines the relationship between $X$ and $Y$.  For each possible index $(x,y)$, $\hat \beta^{-}_{x,y}$ estimates a lower-bound estimate for $\beta_{x,y}$, $\hat \beta_{x,y}$ is a maximum likelihood estimator, and $\hat \beta^{+}$ estimates an upper bound.  Due to identifiability issues $\hat \beta_{x,y}$ is not guaranteed to be consistent.  However, Theorem \ref{thm:consistency} shows that the Markov Link Interval $[\hat \beta^{-}_{x,y},\hat \beta^{+}_{x,y}]$ is asymptotically guaranteed to include the truth.  Above we show all three of these estimators for a selection of values of $(x,y)$; we visualize these values in a grid, where rows correspond to values of $x$, columns correspond to values of $y$, and the color of the square in the $(x,y)$ entry indicates the value of the corresponding parameter.  The values of $\hat \beta$ are nonsensical in many cases.  On the other hand, the Markov Link Interval suggests plausible bounds for the parameters and also indicates further experiments which would help refine these bounds.
}
\end{minipage}\hfill{}
\end{minipage}}\end{figure}

@@

For any cell it is thus impossible to measure a rich variety of cell properties \emph{and} reliably estimate the cell-type.  To understand the relationship between cell-type and these other cell properties, we therefore need some form of unpaired measurement linkage.  This linkage would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.  To facilitate this linkage, one can conduct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.  In each experiment, the probability that a particular cell will be picked out is entirely determined by the cell type $X$, and is thus statistically isolated from the noisy cell-type estimate $Y$.  By varying the experimental parameters we can change the distribution on which cells are picked out.  Letting $L$ denote these corresponding experimental parameters, we see that $L$ can act as a noisy independent indicator about the cell-type.  Markov Link Intervals should therefore give asymptotically consistent bounds on the reliability of the noisy estimate $Y$.

@@

The Markov Link Intervals yield plausible bounds on the true model parameters.  For example, the lower bounds of the Markov Link Interval suggests that a sizable fraction of ``Lamp5 Pdlim5'' cells are indeed classified as ``Lamp5'' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as ``Lamp5 Pdlim5'', but instead only gives coarse class-level cell-type predictions, such as ``Lamp5'').  The upper bounds show that these cells are not commonly categorized into any group besides the `Lamp5' type and the the ``unknown'' type which the Patch method assigns to cells which it cannot identify.  Consider another example: for cells of type ``Sst Myh8-1'',  Markov Link Intervals give us considerably less certainty regarding the related type.  The intervals indicate that the Patch method does not typically classify these cells as ``CR,'' ``L6b,'' ``Meis2,'' ``Sncg,'' or ``Vip'' -- but it does not indicate how they are being classified.  At the extreme end, we have cells of type `Lamp5 Elgn3-1,' for which the Markov Link Intervals are completely vacuous.   It is possible that the Patch method perfectly classifies these ``Lamp5 Elgn3-1' cells; it is possible that it always misclassifies them; the available data cannot tell us either way.

Parameters with large intervals suggest new experiments to perform in order to more closely determine the model parameters.  For example, the significant parameter ambiguity for cells with type ``Lamp5 Egln 1'' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.  If we could find a Cre/Lox procedure that obtained many ``Lamp5 Egln-1'' cells but no cells which Patch categorizes as ``Pvalb,'' this would show that ``Lamp5 Egln 1'' cells are not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells and also selects cells which are largely categorized as `Pvalb' by Patch, this would show the opposite.  If these additional experiments were performed, new Markov Link Intervals could be computed on the new data to determine what aspects of the link are still ambiguous.

@@

\label{sec:conclusions}

We here study ``unpaired measurement linkage,'' the problem of estimating a joint distribution $(X,Y)$ from two separate datasets, one with samples of $X$ and another with samples of $Y$.  Rigorous statistical analysis of this ill-posed problem are complicated by statistical error and identifiability issues.  Here we characterize both issues in the discrete settings when we can observe a third variable $L$ for each sample in both dataset, under the assumption that $L$ and $Y$ are independent conditioned in $X$.

This assumption may hold under a variety of circumstances:

\begin{itemize}
    \item Quality control for manufacturing.   Consider a manufacturing process that produces components; to evaluate production quality, they use two different destructive tests.   The first kind of test is cheap, but its accuracy is unknown.  A more established test is accurate but expensive.  The manufacturer is interested in the link: given results from the cheap test ($X$), what might have been the results from the accurate test ($Y$)?   To estimate the link, we can uses different batches of production ($L$).  For example, $L$ might indicate the day on which a component was made.  The assumption: the link doesn't depend on the day the component was made. This is reasonable if the tests are performed on different days.  In this case, Markov Link Intervals can be used to calibrate the two kinds of tests.

    \item Astronomy photography.  Consider an astronomer taking photographs of stars over time; depending on availability, some pictures are taken using one camera and other pictures are taken using a different camera.  To standardize the photos, the photographer is interested in the link: given a photo from one camera ($X$), what might the picture have looked like if it was taken with the other camera ($Y$)?  To estimate the link, the MLM uses different astronomical targets ($L$).   For example, $L$ might indicate which star is being photographed.  The assumption: the difference between the cameras is the same for each star.  This is reasonable as long as the camera settings are set to the same values for each star.  In this case, Markov Link Intervals can be used to standardize photographs from different cameras.

    \item Personalized medicine.   Consider two sources of knowledge about how transcriptomes might help us personalize cancer medication.  We can measure the transcriptome of cancer cells from inside humans, but experimenting on humans is difficult and morally fraught.  We can also culture immortalized cells from human cancers and measure their transcriptomes;  experimenting on immortalized cells is cheap and straightforward.  To generalize work on immortalized cells to human patients, we need the link: given a cultured cell's transcriptome ($X$), what can we expect about the corresponding in-vivo transcriptome ($Y$)?  To estimate the link, we can use different types of cancer ($L$).  The assumption: the transcriptomic effects of culturing the cells is the same for each cancer type.  This may be reasonable as long as the cancer types are sufficiently similar.  In this case, the Markov Link Intervals  can be used to help generalize results from immortalized cells to real human patients.

    % \item Text/image correspondence.  Consider the problem of automatic captioning.  We can scrape images of people from the internet.  We can also find descriptions of people from literature.  Automatic captioning requires the link: given an image ($X$), what might its text description look like ($Y$)?  To estimate the link, the MLM can use different subject material ($L$).  For example, $L$ might indicate the climate of the location where the photograph was taken.  The assumption: the link is the same for people from different climates.  This is reasonable as long as the photograph itself indicates the important climate information, thereby creating an independence between $Y$ and $L$ after conditioning on $X$.    In this case, the MLM can be used to perform automatic captioning.

    \item Replication crisis and lab effects.  We would like to understand the differences in how two labs perform experiments.  These differences can be understood by looking at the link: given results from one lab ($X$), what results could we expect from another lab ($Y$)?  To estimate the link, we can use different specimen batches ($L$).  For example, $L$ might indicate a batch of mice, half of which were sent to one lab and half of which were sent to another lab.  The assumption: the lab effects are the same for each batch of mice.  
    % This is reasonable as long as the order in which the batches are processed is the same for both labs.  
    In this case, Markov Link Intervals can be used to discover ``lab effects'' that impede successful experimental replication.
\end{itemize}

When the conditional independence assumption holds, Markov Link Intervals  give provably consistent bounds on the joint distribution.  Simulations bear out the consistency of these bounds and also suggest estimation is easiest performs best when $L$ is predictive of $X$ and the relationship between $X$ and $Y$ is either deterministic or completely independent.  An application of the method to cell-type populations demonstrates the method's practical utility for unpaired measurement linkage problems.  One can imagine that these results could be extended to the continuous setting, though additional regularity conditions would be required.

@@

We here give a detailed explanation of how to compute the Markov Link Intervals from two unpaired datasets.

@@

\begin{lemma*}[Unidentifiability]  When $N_L<N_X$ there is no generally consistent estimator to infer $\beta$ from the data $(\DX,\DY)$. 
\end{lemma*}
